A Simple Type Theory for Category Theory

### 2022-04-28 paper/paper.tex
missing last step of interchange
complete draft of the semantics section

In terms of the syntax of SCT, the category $C_0$ models the category
and functor judgments, the category $C_1$ models the profunctor
\emph{contexts} and substitutions and the category $C_d$ models the
profunctor judgment. The profunctor $C_\vdash$ models the
transformation judgment and the functor $i$ models the construction of
a ``singleton'' context $\alpha:\cat C,x:R,\beta:\cat D$.

@@

Next, each of the connectives of SCT are modeled by different
universal properties that are profunctor isomorphisms involving
principally $C_\vdash$.
TODO: write them down.

@@

\begin{definition}
  A Category with Families is...
\end{definition}

\begin{definition}
  A hyperdoctrine of equipment with displays (HED) is a pair of a
  category with families $Ctx$ modeling $\Sigma,\Pi,\textrm{Id}$ and a
  DCD internal to $\Set^{Ctx^o}$ with categories, functors, displays
  and unary cells representable as types.
\end{definition}

Again this notion is straightforwardly, if laboriously, interpretable
as a GAT.

%% \begin{definition}[Unit]
%%   A unit $U_{\cat C}\in V_h$ with $d^\pm U_{\cat C} = \cat C$ and an op-cartesian 2-cell
%%   \[ r : V_h(\cat C, U_{\cat C}) \]
%%   with $d^\pm r = \id_{\cat C}$.
%% \end{definition}

%% \begin{definition}[Tensor]
%%   If $R,Q \in V_h$ with $d^+R = d^-Q$, a tensor is a horizontal arrow
%%   $R \odot Q \in V_h$ with $d^+ (R\odot Q) = d^+ R$ and $d^-(R \odot
%%   Q) = d^- Q$ with an op-cartesian 2-cell
%%   \[ p : V_h(R,Q;R \odot Q) \]
%% \end{definition}

\begin{definition}{Indexed Virtual Equipment}
  A VE-hyperdoctrine is a pair of a category with families $(C,
  \textrm{Ty}, \textrm{Tm})$ and a functor $\cat C^{op} \to VE$. A
  morphism of VE-hyperdoctrines
  \[(F,\phi) : (\cat C, \mathcal V : \cat C^{o} \to VE) \to (\cat D, \mathcal U : \cat D^{o} \to VE)\]
  is a pair of a functor of categories with families $F : \cat C \to
  \cat D$ and a natural transformation $\phi : \mathcal V
  \Rightarrow\mathcal U \circ F^{o}$.
\end{definition}

@@

\subsection{Generalized Algebraic Theory of VE Hyperdoctrines}

To simplify our initiality proof, we use a \emph{generalized
algebraic} presentation of our notion of
model\cite{cartmell,sterling-eat}. To describe this, we give a
``combinator'' syntax for VE hyperdoctrines, where we have explicit
sorts for contexts $\Gamma$ and $\Phi$ and explicit judgments for
substitutions $\gamma : \Delta \to \Gamma$ and $\phi : \Psi
\vdash_{f,g} \Phi$.
%
While the explicit description of the theory is quite long, we can
describe it in simple modular pieces.
%
First, we use the standard theory of a category with families, which
describes the algebraic structure of contexts $\Gamma$, types $\Gamma
\vdash A$ and terms $\Gamma \vdash M : A$.
%
Second, we add to that a theory for a slight generalization of virtual
double categories we call \emph{double categories with display
pro-arrows}.
%
The difference is that in a double category with display pro-arrows,
we have a true double category with strict vertical and horizontal
composition, among whom are the ``display'' pro-arrows.
%
The display pro-arrows model the profunctors of the syntax, which
don't have a strict composition, whereas the arbitrary pro-arrows
model the contexts $\Phi$, which do have a strict composition, given
by context concatenation $\Phi,\Psi$.
%
The difference is that in a virtual double category the contexts are
assumed to be given inductively by formal compositions of the display
pro-arrows, whereas in models of this algebraic theory, the contexts
might include other exotic objects.
%
This difference is analogous to the difference between a category with
families and a contextual category in models of dependent type theory,
and we fully expect that there is a similar relationship between the
models: virtual double categories should form a co-reflective
subcategory characterized as those double categories with display
pro-arrows where all pro-arrows are equal to some concatenation of
displays.
%
Finally, we parameterize all of the constructions of the theory of
double categories with pro-arrows by contexts $\Gamma$ and add in
equations that substitution commutes with all double category
operations.
%
This means the theory presents a category with families with an double
category with display pro-arrows internal to the category of
presheaves on the category of contexts, i.e., a hyperdoctrine of
double categories with display pro-arrows.

The main advantage of this algebraic approach is that the initiality
theorem for the combinatory syntax is trivial. Instead, the work is to
show that our type theoretic syntax, which restricts combinatory rules
to make substitution an admissible rather than primitive operation, is
equivalent to the combinatory syntax.

\begin{figure}
  \begin{mathpar}
    \inferrule
    {}
    {\Gamma : \algCtx}

    \inferrule
    {\Delta : \algCtx \and \Gamma : \algCtx}
    {\gamma : \algSubst(\Delta,\Gamma)}

    \inferrule
    {\Gamma : \algCtx}
    {C : \algCat(\Gamma)}

    \inferrule
    {\Gamma : \algCtx \and C : \algCat(\Gamma) \and D : \algCat(\Gamma) \and
    }
    {F : \algVarr(\Gamma,C,D)}

    \inferrule
    {\Gamma : \algCtx \and C : \algCat(\Gamma) \and D : \algCat(\Gamma) \and
    }
    {P : \textrm{Profunctor}(\Gamma,C,D)}

    \inferrule
    {\Gamma : \algCtx \and C : \algCat(\Gamma) \and D : \algCat(\Gamma)
    }
    {P : \textrm{ProfCtx}(\Gamma,C,D)}

    \inferrule
    {\Gamma : \textrm{Ctx}\and
      C  : \algCat(\Gamma) \and D  : \algCat(\Gamma) \and
      C' : \algCat(\Gamma) \and D' : \algCat(\Gamma) \and
      f : \algVarr(\Gamma,C,C') \and g : \algCat(\Gamma,D,D') \and
      \Phi : \textrm{ProfCtx}(\Gamma,C,D)\and
      \Psi : \textrm{ProfCtx}(\Gamma,C',D')
    }
    {\phi : \textrm{ProfSubst}(\Gamma,C,D,C',D',f,g,\Phi,\Psi)}

    \inferrule
    {\Gamma : \textrm{Ctx}\and
      C  : \algCat(\Gamma) \and D  : \algCat(\Gamma) \and
      C' : \algCat(\Gamma) \and D' : \algCat(\Gamma) \and
      f : \algVarr(\Gamma,C,C') \and g : \algCat(\Gamma,D,D') \and
      \Phi : \textrm{ProfCtx}(\Gamma,C,D)\and
      P : \textrm{ProfCtx}(\Gamma,C',D')
    }
    {t : \textrm{Trans}(\Gamma,C,D,C',D',f,g,\Phi,P)}
  \end{mathpar}
  \caption{Sorts of Algebraic SCT}
\end{figure}

\begin{figure}
  \begin{mathpar}
    \cdots
  \end{mathpar}
  \caption{Algebraic Dependent Type Theory}
\end{figure}

The algebraic theory of a double category with displays indexed by a
category is given in Figure~\ref{fig:alg-dc-disp}.
%
First, we give the categories and functors the theory of a category
indexed by a category.
%
Next, we give categories and \emph{profunctor contexts} the same
theory, where the identity $\id_h$ is the ``empty'' context $\alpha:C$ and the composition $\Phi \circ_h$.
%

TODO: include all of the interactions with substitutions and restrictions.

\begin{figure}
  \begin{mathpar}
    % theory of an indexed category (vertical)
    \inferrule
    {g : \algVarr(D,E) \and \algVarr(C,D)}
    {g \circ_v f : \algVarr(C,E)}

    (g \circ_v f) \circ_{vs} \gamma = (g \circ_{vs} \gamma) \circ_v (f \circ_s \gamma)

    (h \circ_v g) \circ_v f = h \circ_v (g \circ_v f)

    \inferrule
    {}
    {\id_v : \algVarr(C,C)}

    \id_v \circ_{vs} \gamma = \id_h

    (f \circ_v \id_v) = f\and \id_v \circ_v f = f

    % theory of an indexed category (horizontal)

    \inferrule
    {\Psi : \algHCtx(D,E) \and \algHCtx(C,D)}
    {\Psi \circ_h \Phi : \algHCtx(C,E)}

    (\Psi \circ_h \Phi) \circ_{hs} \gamma = (\Psi \circ_{hs} \gamma) \circ_h (\Phi \circ_s \gamma)

    (\Sigma \circ_h \Psi) \circ_h \Phi = \Sigma \circ_h (\Psi \circ_h \Phi)

    \inferrule
    {}
    {\id_h : \algHCtx(C,C)}

    \id_h \circ_{hs} \gamma = \id_h

    (\Phi \circ_h \id_h) = \Phi\and \id_h \circ_h \Phi = \Phi

    % embedding of displays

    \inferrule{P : \algHarr(C,D)}{i(P) : \algHCtx(C,D)}

    % 2-cells: substitutions
    % vertical category
    % horizontal category
    % interchange law
    \inferrule
    {\psi : \algHSubst(\Phi, f',g',\Psi)\and \phi : \algHSubst(\Sigma,f,g,\Phi)
    }
    {\psi \circ_{2v} \phi : \algHSubst(\Sigma, f' \circ f, g' \circ g, \Psi)}

    \inferrule
    {}
    {\id_{2v} : \algHSubst(\Phi,\id,\id,\Phi)}

    (\phi \circ_{2v} \psi) \circ_{2v} \sigma = \phi \circ_{2v} (\psi \circ_{2v} \sigma)\\
    \phi \circ_{2v} \id_{2v} = \phi \and \phi = \id_{2v} \circ_{2v} \phi\\

    \inferrule
    {\phi_1 : \algHSubst(\Psi_1,f,g,\Phi_1)\and \phi_2 : \algHSubst(\Psi_2,f,g,\Phi_2)
    }
    {\phi_1 \circ_{2h} \phi_2 : \algHSubst(\Psi_1 \circ_h \Psi_2, f, h, \Phi_1 \circ_h \Phi_2)}

    \inferrule
    {}
    {\id_{2h} : \algHSubst(\id_h,f,f,\id_h)}

    (\phi \circ_{2h} \psi) \circ_{2h} \sigma = \phi \circ_{2h} (\psi \circ_{2h} \sigma)\\
    \phi \circ_{2h} \id_{2h} = \phi \and \id_{2h} \circ_{2h} \phi = \phi\\

    (\phi_1 \circ_{2h} \phi_2) \circ_{2v} (\psi_1 \circ_{2h} \psi_2) = (\phi_1 \circ_{2v} \psi_1) \circ_{2h} (\phi_2 \circ_{2v} \psi_2)

  \end{mathpar}
  \caption{Double Category with Display Pro-arrows}
  \label{fig:alg-dc-disp}
\end{figure}

\begin{figure}
  \begin{mathpar}
    % Restrictions
    \inferrule
    {P : \algHarr(\Gamma,C',D')
    \and f : \algVarr(\Gamma,C,C')\and
    \and g : \algVarr(\Gamma,D,D')}
    {P \circ_r (f,g) : \algHarr(\Gamma,C,D)}

    (P \circ_r (f,g) \circ_s \gamma) = (P \circ_s \gamma) \circ_r (f \circ_s \gamma, g \circ_s \gamma)

    P \circ_r (f,g) \circ_r (f',g') = P \circ_r (f \circ_r f', g \circ_r g')

    P \circ_r (\id,\id) = P

    % Prof Multiplicatives
    % LHom
    \inferrule
    {P : \algHarr(C, D) \and Q : \algHarr(C,E)}
    {\algLHom(P,Q) : \algHarr(D,E)}

    \algLHom(P,Q) \circ_{hs} \gamma = \algLHom(P \circ_{hs} \gamma, Q \circ_{hs} \gamma)\and
    \algLHom(P,Q) \circ_{r} (f,g) = \algLHom(P \circ_r (\id,f), Q \circ_r(\id,g))

    \inferrule
    {}
    {\algLHomE(P,Q) : \algHSubst(i(P)\circ_h i(\algLHom(P,Q)),\id_v,\id_v,i(Q))}

    \inferrule
    {t : \algHSubst(i(P) \circ_h \Phi, \id,g, i(Q))}
    {\algLHomI(\Phi,g,t,P,Q) : \algHSubst(\Phi,\id,g,i(\algLHom(P,Q)))}

    \algLHomE \circ_{2v} (\id_v(P) \circ_h \algLHomI(t)) = t\and
    \theta = \algLHomI(\algLHomE \circ_{2v} (\id_v(P) \circ_h \theta))\and

    % RHom
    \inferrule
    {P : \algHarr(C, E) \and Q : \algHarr(D,E)}
    {\algRHom(P,Q) : \algHarr(C, D)}

    \algRHom(P,Q) \circ_{hs} \gamma = \algRHom(P \circ_{hs} \gamma, Q \circ_{hs} \gamma)\and
    \algRHom(P,Q) \circ_{r} (f,g) = \algRHom(P \circ_r (f, \id), Q \circ_r(g,\id))

    \inferrule
    {}
    {\algRHomE(P,Q) : \algHSubst(i(\algRHom(P,Q)) \id_h i(Q),\id_v,\id_v,i(P))}

    \inferrule
    {t : \algHSubst(\Phi \circ_h i(Q), f, \id, i(P))}
    {\algRHomI(\Phi,f,t,P,Q) : \algHSubst(\Phi,\id,g,i(\algLHom(P,Q)))}

    \algRHomE \circ_{2v} (\algRHomI(t) \circ_h \id(Q)) = t\and
    \theta = \algRHomI(\algRHomE \circ_{2v} (\theta \circ_h \id(Q)))\and

    % Unit
    \inferrule
    {C : \algCat}
    {\algUnit(C) : \algHarr(C,C)}

    \algUnit(C) \circ_{hs} \gamma = \algUnit(C \circ_{cs} \gamma)

    \inferrule
    {}
    {\algUnitI(C) : \algSubst(\id_h(C),\id,\id,i(\algUnit(C)))}

    \inferrule
    {t : \algHSubst(\Phi \id_h \Psi,f,g,i(R))}
    {\algUnitE(t) : \algHSubst((\Phi \circ_h i(\algUnit(C)) \circ_h \Psi), f, g, i(R))}

    \algUnitE(t) \circ_{2v} (\id_{2v}(\Phi) \circ_h \algUnitI(C) \circ_h \id_{2v}(\Psi)) = t\and
    s = \algUnitE(s \circ_{2v} (\id_{2v}(\Phi) \circ_h \algUnitI(C) \circ_h \id_{2v}(\Psi)))

    % Tensor
    \inferrule
    {P : \algHarr(C,D) \and Q : \algHarr(D,E)}
    {\algTensor(P,Q) : \algHarr(C,E)}

    \algTensor(P,Q) \circ_{hs} \gamma = \algTensor(P \circ_{hs} \gamma, Q \circ_{hs} \gamma)\and
    \algTensor(P,Q) \circ_{r} (f,g) = \algTensor(P \circ_r (f,\id), Q \circ_r (\id,g))

    \algTensorI(P,Q) : \algHSubst(i(P) \circ_h i(Q),\id,\id,i(\algTensor(P,Q)))

    \inferrule
    {t : \algHSubst(\Phi \circ_h i(P) \circ_h i(Q) \circ_h \Psi, f,g,i(R))}
    {\algTensorE(t) : \algHSubst(\Phi \circ_h i(\algTensor(P,Q)) \circ_h \Psi, f,g,i(R))}

    \algTensorE(t) \circ_{2v} (\id_{2v}(\Phi) \circ_{2h} \algTensorI(P,Q) \circ_{2h} \id_{2v}(\Psi)) = t\and
    s = \algTensorE(s \circ_{2v} (\id_{2v}(\Phi) \circ_{2h} \algTensorI(P,Q) \circ_{2h} \id_{2v}(\Psi)))

    % Cat Multiplicatives
    % Cat of Elements
    \inferrule
    {P : \algHarr(\Gamma,C,D)}
    {\algElts(C,D,P) : \algCat(\Gamma)}

    \inferrule
    {}
    {\pi^- : \algVarr(\algElts(C,D,P),C)}

    \inferrule
    {}
    {\pi^+ : \algVarr(\algElts(C,D,P),D)}

    \inferrule
    {}
    {\pi^p : \algHSubst(\id_h(\algElts(C,D,P)),\pi^-,\pi^+,P)}

    \inferrule
    {f : \algVarr(E,C) \and g : \algVarr(E,D)\and
      t : \algHSubst(\id_h(E), f, g, P)
    }
    {\algEltsI(t,f,g) : \algVarr(E,\algElts(C,D,P))}

    \pi^- \circ_v \algEltsI(t,f,g) = f\and
    \pi^+ \circ_v \algEltsI(t,f,g) = g\and
    \pi^p \circ_{2v} \algEltsI(t,f,g) = \alpha

    \inferrule
    {h : \algVarr(E,\algElts(C,D,P))}
    {h = \algEltsI(\pi^-\circ_v h,\pi^+\circ_vh,\pi^p \circ_{2v} \id_{2h}(h))}

    % Presheaf cats

    % Cat Additives
    % Product
    % Terminal

    % Prof Additives
    % Product
    % Terminal
  \end{mathpar}
  \caption{Multiplicatives}
\end{figure}

\begin{figure}
  \begin{mathpar}
    % Internalization
    % Cats
    \inferrule
    {}
    {\algCatTy : \algTy(1)}

    \inferrule
    {C : \algCat(\Gamma)}
    {\algCatQt(C) : \algTm(\Gamma,\algCatTy \circ !)}

    \algCatQt(C) \circ_{tms} \gamma = \algCatQt(C \circ_{cs} \gamma)\and

    \inferrule
    {M : \algTm(\Gamma,\algCatTy \circ !)}
    {\algCatUnqt(M) : \algCat(\Gamma)}

    \algCatUnqt(M) \circ_{cs} \gamma = \algCatUnqt(M \circ_{tms} \gamma)\and
    \algCatUnqt(\algCatQt(C)) = C\and
    M = \algCatQt(\algCatUnqt(M))

    % Functor
    % Prof
    % ProfCtx?
    % Trans
    \inferrule
    {P : \algHarr(\Gamma,C,C)}
    {\algTransTy(P) : \algTy(\Gamma)}

    \inferrule
    {t : \algHSubst(\id_h,\id_v,\id_v,i(P))}
    {\algTransQt(t) : \algTm(\Gamma,\algTransTy(P))}

    \algTransQt(t) \circ_{tms} \gamma = \algTransQt(t \circ_{2s} \gamma)

    \inferrule
    {}
    {\algTransUnqt : \algHSubst(ext(\Gamma,\algTransTy(P)),\id_h,\id_v,\id_v,i(P \circ_{hs} \pi_1))}

    \algTransUnqt \circ_{2s} (\gamma,\algTransQt(t)) = t\\

    % M : Tm(G,∀ α. P α α)
    % unquote : Subst(G,t : ∀ α:C. P α α; α:C, id,id, P α α)
    % M = unquote()
    M = \algTransQt(\algTransUnqt \circ_{2s} ext_s(\id,M))
  \end{mathpar}
  \caption{Internalization}
\end{figure}

\subsection{Equivalence between Natural Deduction and Combinatory Syntaxes}

Now we aim to prove that our natural deduction syntax of SCT is sound
and complete for the algebraic notion of hyperdoctines of double
categories with displays.
%
In category-theoretic terms, we want to show that the syntax of SCT
presents an \emph{initial object} in the category of models.
%
This is proved in two steps: first, we show that SCT is a model

\begin{construction}{Completeness}
  SCT presents a hyperdoctrine of double categories with displays.
\end{construction}
\begin{proof}
  First, the sorts.
  \begin{align*}
    \eta(\algCtx) &= \{ \Gamma | \Gamma \isadtctx \}\\
    \eta(\algSubst(\Delta,\Gamma)) &= \{ \gamma | \eta(\Delta) \vdash \gamma : \eta(\Gamma) \}\\
    \eta(\algTy(\Gamma)) &= \{ A |  \eta(\Gamma) \vdash A \isaTy \}\\
    \eta(\algTm(\Gamma,A)) &= \{ M | \eta(\Gamma) \vdash M : \eta(A) \}\\
    \eta(\algCat(\Gamma)) &= \{ \cat C | \eta(\Gamma) \vdash \cat C \isaCat \}\\
    \eta(\algVarr(\Gamma,C,D)) &= \{ a | \eta(\Gamma)|\alpha:\eta(C) \vdash a : \eta(D) \}\\
    \eta(\algHarr(\Gamma,C,D)) &= \{ R | \eta(\Gamma)|\alpha:\eta(C);\beta:\eta(D) \vdash R \isaSet \}\\
    \eta(\algHCtx(\Gamma,C,D)) &= \{ \Phi | \eta(\Gamma) \vdash \Phi \isavectx \wedge d^-(\Phi) = \eta(C) \and d^+(\Phi) = \eta(D) \}\\
    \eta(\algHSubst(\Gamma, \Psi, f, g, \Phi)) &= \{ \phi | \eta(\gamma)\pipe \eta(\Psi) \vdash \phi : \eta(\Phi) \wedge d^-\phi = \eta(f) \wedge d^+\phi = \eta(g) \}\\
    1 &= \cdot\\
    ! &= \cdot\\
    ext(\Gamma,A) = \Gamma, x:A \\
    \cdots\\
    \Phi \circ_h \Psi &= \Phi \jnctx \Psi\\
    \id_h(\cat C) &= \alpha:\cat C\\
    i(\cat C, \cat D, R) &= \alpha:\cat C, x: R, \beta:\cat D\\
    \phi \circ_{2v} \psi &= \phi[\psi]\\
    \id_{2v}(\Phi) &= \id_\Phi\\
    \id_{2h}(\alpha:\cat C \vdash a : \cat D) &= a/\beta\\
    \phi \circ_{2h} \psi &= \phi \jnctx \psi
  \end{align*}
\end{proof}
\begin{corollary}
  There is a unique homomorphism of HDCDs
  \[ \eta : Comb \to SCT \]
\end{corollary}

\begin{construction}{Soundness}
  We construct a homomorphism of HDCDs
  \[ i : SCT \to Comb \]
  such that the composite
  \[ \eta \circ i : SCT \to SCT \]
  is equal to the identity.
\end{construction}
\begin{proof}
  \begin{align*}
    i(\cdot) &= 1\\
    i(\Gamma,x:A)  &= ext(i(\Gamma),ext(A))\\
    \cdots
    i(\Gamma \pipe \alpha:\cat C \vdash \alpha:\cat C) &= \id_v\\
    i(\Gamma \pipe \alpha:\cat C \vdash M a :\cat C) &= (\algVarrUnqt \circ_{vs} (\id, (i(M)))) \circ_v i(a)\\
    % Presheaf intro
    % elements intro
    i(\Gamma \pipe \alpha:\cat C \vdash (a,t,b) : \Sigma_{\alpha:\cat C;\beta:\cat D} R) &= \algEltsI(i(a),i(t),i(b))\\
    %
  \end{align*}
\end{proof}

\begin{corollary}
  SCT is initial in the category of HDCDs.
\end{corollary}
\begin{proof}
  Since Comb is initial, the composite $i \circ \eta$ is the identity,
  and as above the composite $\eta \circ i$ so SCT is isomorphic to
  Comb, and therefore it is also initial.
\end{proof}

@@

Our main intended models are when the indexing category is $\Set$ and
the hyperdoctrine is given by families of objects, arrows etc, of a
fixed virtual equipment $\mathcal V$, indexed by the set
$\sem\Gamma$. We can formalize this as the ``internal logic'' of the
virtual equipment $\mathcal V$.

\begin{definition}[Powers of Virtual Equipments]
  Let $\mathcal V$ be a virtual equipment and let $X$ be a set.
  Define a virtual equipment, the ``power'' $\mathcal V^X$ as follows.

  \begin{enumerate}
  \item Objects are functions $X \to \mathcal V_0$.
  \item Vertical morphisms $\mathcal V^X_v(A,B)$ are functions
    $\Pi_{x\in X} V_v(A x,B x)$. With pointwise composition, identity.
  \item Horizontal morphisms $\mathcal V^X_h$ are functions $\Gamma \to \mathcal V_h$, with
    $(d^\pm R)x = d^\pm (Rx)$.
  \item Restriction of $R$ along $f,g$ is given by $(R(f,g))x = (R x)(f x,g x)$.
  \item A 2-cell from $\Phi$ to $R$ is a family $\Pi_{x\in X} \mathcal
    V^X_h(\Phi x,R x)$ with composition given pointwise.
  \end{enumerate}
\end{definition}

\begin{definition}[Internal Logic]
  Given any virtual equipment $\mathcal V$, we can construct the
  ``internal logic'' by indexing $\mathcal V$ by $\Set$:

  \begin{enumerate}
  \item The indexing cateogry is $\Set$.
  \item The equipment is defined as $\mathcal U\Gamma = \mathcal V^\Gamma$.
  \item Restriction by a function $\gamma : \Delta \to \Gamma$ is
    given pointwise.
  \end{enumerate}
\end{definition}

%% \section{Conservativity of Homs, Tensors}

>>>>

In terms of the syntax of SCT, the category $C_0$ models the category
and functor judgments, the category $C_1$ models the profunctor
\emph{contexts} and substitutions and the category $C_d$ models the
profunctor judgment. The profunctor $C_\vdash$ models the
transformation judgment and the functor $i$ models the construction of
a ``singleton'' context $\alpha:\cat C,x:R,\beta:\cat D$. The fact
that $i$ is fully faithful corresponds to the fact that we don't
provide in the syntax a notion of morphism of profunctors separate
from the notion of substitution of singleton contexts.

@@

The final notion of composition that we require for our structure is
that display morphisms have \emph{restrictions} along vertical arrows.

\begin{definition}
  A DCD is an Equipment with displays when
  \begin{enumerate}
  \item The functor $(s,t) : C_d \to C_0$ is a split fibration, with
    restriction operation written as $(f,g)^*R$.
  \item The profunctor $C_\vdash$ respects restrictions in that
    \[ C_\vdash^{f',g'}(\Phi, (f,g)^*R) \cong C_\vdash^{f\circ f',g\circ g'}(\Phi, R) \]
  \end{enumerate}
\end{definition}
Specializing this property give us that $C_\vdash^{f,g}(\Phi,R) \cong
C_\vdash^{\id,\id}(\Phi,(f,g)^*R)$, which justifies that we only need
to give syntax for terms $\Phi \vdash t : R$ where $\Phi$ and $R$ have
the same covariant/contravariant dependencies.

We can then define the remaining connectives by UMPs in an equipment
with displays, adding in the extra strictness criterion that these
connectives appropriately commute with restriction.

\begin{figure}
  \begin{mathpar}
    C_\vdash^{f,g}(\Phi \circ_{C} \Psi, R) \cong C_\vdash^{f,g}(\Phi \circ_C i(U_C) \circ_c \Psi, R) \\
    C_\vdash^{f,g}(\Phi \circ_{C} iP \circ_D iQ \circ_E \Psi, R) \cong C_\vdash^{f,g}(\Phi \circ_C i(P \odot Q) \circ_E \Psi, R) \and (f,g)^*(P \odot Q) = (f,\id)^*P \odot (\id,g)^*Q\\
    C_\vdash^{\id_C,g}(P \circ_c \Phi, R) \cong C_\vdash^{\id_D,g}(\Phi, P \triangleright Q) \and (f,g)^*(P \triangleright Q) = (\id,f)^*P \triangleright (\id,g)^*Q\\
    C_\vdash^{f, \id_C}(\Phi \circ_c R, P) \cong C_\vdash^{f, \id_D}(\Phi, P \triangleleft Q)\and (f,g)^*(P \triangleleft Q) = (f,\id)^*P \triangleleft (g,\id)^*Q\\
    C_\vdash^{f,g}(\Phi,1) \cong 1 \and (f,g)^*1 = 1\\
    C_\vdash^{f,g}(\Phi,P \times Q) \cong C_\vdash^{f,g}(\Phi,P) \times C_\vdash^{f,g}(\Phi, Q)\and (f,g)^*(P \times Q) = (f,g)^*P \times (f,g)^*Q\\
    C_0(C,\smallint R) \cong C_\vdash(\id_C, R)\\
  \end{mathpar}
  \caption{UMPs in an Equipment with Displays}
\end{figure}

Next, each of the connectives of SCT are modeled by different
universal properties that are profunctor isomorphisms involving
principally $C_\vdash$. We define these in Figure\ref{fig:eq-umps}

@@

\begin{definition}
  A Category with Families consists of
  \begin{enumerate}
  \item A category $\textrm{Ctx}$
  \item with a terminal object $1 \in \textrm{Ctx}$
  \item A presheaf $\textrm{Ty} : \Set^\textrm{Ctx}$ of types in context.
  \item A presheaf $\textrm{Tm} : \Set^{\int \textrm{Tm}}$ of terms of
    a given type.
  \item With a context extension functor $\textrm{ext} : \int
    \textrm{Ty} \to \textrm{Ctx}$ with universal property:
    \[ \textrm{Ctx}(\Delta,\textrm{ext}(A)) \cong (\gamma : \textrm{Ctx}(\Delta,\textrm{ctx}(A))) \times \textrm{Tm}(A[\gamma])  \]
  \end{enumerate}
\end{definition}

This can be extended in a standard way to model $\Pi,\Sigma, \Id$
types \citet{cwf-stuff}.
%
The final connectives we need are the ``universes'' that internalize
the double category structure as types in the CwF.

\begin{definition}
  Let $(\textrm{Ctx},\textrm{Ty},\textrm{Tm},\textrm{ext})$ be a
  category with families and let $\mathcal V$ be a DCD internal to
  $\Set^\textrm{Ctx}$. Then
  \begin{enumerate}
  \item $\mathcal V$ objects are representable as types if there is a
    type given by a section $\textrm{Ob} \in \textrm{Ty}(1)$ with a
    natural isomorphism
    \[ \textrm{Tm}(!_\Gamma^*\textrm{Ob}) \cong \textrm{Ob}(\mathcal V)_\Gamma \]
  \item $\mathcal V$ vertical arrows are representable as types if
    there is a type
    \[ \textrm{Tm}(((\gamma,\lceil a\rceil),\lceil b\rceil)^*\textrm{Functor} \cong
    \textrm{VArr}(\mathcal V_{\Delta})(a,b)
    \]
    and similarly horizontal displays are representable as types if
    \[ \textrm{Tm}(((\gamma,\lceil a\rceil),\lceil b\rceil)^*\textrm{Profunctor} \cong
    \textrm{Disp}(\mathcal V_{\Delta})(a,b)
    \]
  \item Closed $\mathcal V$ transformations are representable as types if
    \[ \textrm{Tm}() \]
  \end{enumerate}
\end{definition}

We then arrive at our final notion of model by collecting all of this
algebraic structure into one definition.
\begin{definition}
  A hyperdoctrine of equipments with displays (HED) is a pair of a
  category with families $Ctx$ modeling $\Sigma,\Pi,\Id$ and a DCD
  internal to $\Set^{Ctx^o}$ with categories, functors, displays and
  unary cells representable as types.
\end{definition}

This notion of HED is straightforwardly, if laboriously, interpretable
as a GAT.

@@

\subsection{Initiality}

In this section we give a sketch of an initiality proof of the syntax
of SCT in HEDs. Note that we consider all the following sets up to
$\alpha$-equivalence of names.

\begin{construction}{Completeness}
  SCT presents a HED with
  \begin{enumerate}
  \item Category with Families provided by the contexts,
    substitutions, types and terms.
  \item Objects of $C_0^\Gamma$ provided by the terms $\Gamma \vdash \cat C \isaCat$
  \item Morphisms of $C_0^\Gamma$ provided by terms $\Gamma
    \alpha:\cat C \vdash b \cat D$ with source $\cat C$ and target
    $\cat D$. With identity given by a variable and composition by identity.
  \item Objects of $C_1^\Gamma$ are provided by the profunctor
    contexts $\Gamma \vdash \Phi \isavectx$ with morphisms given by
    transformation substitutions $\Gamma \pipe \Phi \vdash \psi :
    \Psi$.
  \item Horizontal composition of contexts is defined by gluing along
    a variable $\Phi\jnctx \Psi$ and the horizontal identity is given
    by the single category variable $\alpha:\cat C$.
  \item Objects of $C_\vdash^\Gamma$ are provided by the profunctors
    $\Gamma\alpha:\cat C;\beta:\cat D \vdash R$, and a morphism from
    $\alpha:\cat C;\beta:\cat D \vdash R$ to $\alpha':\cat
    C';\beta':\cat D' \vdash R'$ is a triple $(a,t,b)$ of functor
    terms $\alpha:\cat C \vdash a : \cat C'$ and $\beta:\cat D\vdash b
    : \cat D'$ and a term $\alpha:\cat C,x:R,\beta:\cat D \vdash t :
    R'[a/\alpha';b/\beta']$

    Composition $(a,t,b) \circ (a',t',b') = (a[a'],t[t'],b[b'])$ is
    given by component-wise single-place substitution with the triple
    of all variables the identity.
  \item The inclusion functor $i^\Gamma : C_d \to C_1$ is defined on
    types as the singleton context $i(R) = \alpha:\cat
    C,x:R,\beta:\cat D$ and on morphisms as the singleton
    substitution.
  \item The restriction operation $R \circ (a,b)$ is defined as the
    substitution $R[a;b]$.
  \item The remaining universal properties are satisfied using the
    analogous type theoretic construction.
  \end{enumerate}
\end{construction}

The associativity, unit and interchange laws of a category are proven
in the \cref{appendix}.

\begin{theorem}{(Sketch) Soundness/Initiality}
  Given any HED $\mathcal V$, there is a unique homomorphism of HEDs $\sem\cdot$
  from $\textrm{Syn}(\textrm{SCT})$ to $\mathcal V$.
\end{theorem}
\begin{proof}
  Roughly the converse of the completeness theorem. For instance,
  profunctor contexts are interpreted using horizontal identity and
  composition as well as the $i$ functor. Preservation of the
  restriction operation requires that restriction commutes with the
  display connectives.
\end{proof}

Details of the interpretation of profunctor contexts and substitutions
are given in the appendix\cref{appendix}.

@@

We now describe how to interpret the syntax in any virtual equipment
(that can model the connectives). First, from any virtual equipment we
can construct a DCD:

\begin{definition}
  Given a virtual equipment $\mathcal V$, define a DCD by taking the
  horizontal morphisms of $\mathcal V$ as the display morphisms and
  the horizontal morphisms as the ``composable strings'' of horizontal
  morphisms. Then we define $C_\vdash$ using the notion of
  ``multi-2-cell''.
\end{definition}

Next, we can upgrade any DCD in $\Set$ to a DCD internal to
$\Set^\Set$ by taking the presheaf of objects, vertical arrows, etc as
families of objects, vertical arrows etc indexed by the indexing set
$\Gamma$.
\begin{definition}[Powers of Virtual Equipments]
  Let $\mathcal V$ be a virtual equipment and let $X$ be a set.
  Define a virtual equipment, the ``power'' $\mathcal V^X$ as follows.

  \begin{enumerate}
  \item Objects are functions $X \to \mathcal V_0$.
  \item Vertical morphisms $\mathcal V^X_0(A,B)$ are functions
    $\Pi_{x\in X} V_0(A x,B x)$. With pointwise composition, identity.
  \item Horizontal morphisms $\mathcal V^X_1$ are functions $\Gamma
    \to \mathcal V_1$.
  \item A 2-cell from $\Phi$ to $R$ is a family $\Pi_{x\in X} \mathcal
    V^X_h(\Phi x,R x)$.
  \end{enumerate}
  All operations are simply lifted pointwise.
\end{definition}

This defines the ``natural'' interpretation of SCT in a virtual
equipment once we see that the powering construction is functorial in
$\Gamma$.
\begin{definition}[Internal Logic]
  Given any virtual equipment $\mathcal V$, we can construct the
  ``internal logic'' by indexing $\mathcal V$ by $\Set$:
